Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-23 setup.tex
introduce alias for Omega_K,  \OmK, think I got all...

\subsection{Thermal relaxation}
We model radiative effects as thermal
relaxation with a timescale $t_c$, i.e.,
\begin{align}\label{thermal_relax}
  \Lambda  \equiv \frac{\rho\mathcal{R}}{\mu}\frac{T - T_0}{t_c},
\end{align}
where $T_0=T_0(r,z)$ is the equilibrium temperature.  We 
define the dimensionless cooling time 
\begin{align}\label{beta_def}
  \beta \equiv \Omega_k t_c
\end{align}
relative to the Keplerian frequency, $\Omega_k=\sqrt{GM_*/r^3}$.

@@

We consider disk equilibria in which the pressure $P$ and density
$\rho$ are related by   
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a  constant parameter  and the
function $c_{s0}(r)$ is prescribed below. Choosing $\Gamma=1$ gives  a 
locally vertically isothermal disk and $c_{s0}(r)$ is the mid-plane
sound-speed. 
It shall be convenient to define the function $c_s(r,z)$ 
\begin{align}
  c_s^2\equiv % \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
  \frac{\Gamma P}{\rho}, 
\end{align}
which is related to the isothermal sound-speed $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$. We take $c_{s0}(r)$ to be another power-law:  
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index, and
define   
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k$ is the Keplerian azimuthal velocity and
$H$ is a characteristic scale-height.  Thus, $\epsilon$
is also a characteristic disk aspect-ratio. We are 
interested in thin disks with $\epsilon \ll 1$.%,%  as relevant for
% protoplanetary disks \citep{chiang97}.

@@

Similarly, the equilibrium rotation is 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}=1 +
  \left(p+\frac{s}{\Gamma}\right)\epsilon^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
where $s\equiv q+p(1-\Gamma)$. We also define % the radial shear rate  
% \begin{align}
%   S(r,z) \equiv r\frac{\p\Omega}{\p r},  
% \end{align}
% and
the square of the epicyclic frequency $\kappa$ 
\begin{align}
  \kappa^2(r,z) \equiv \frac{1}{r^3}\frac{\p }{\p
    r}\left(r^4\Omega^2\right). %2\Omega(2\Omega + S). 
\end{align}

\subsubsection{Vertically isothermal disks}
For $\Gamma=1$ the equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ are given by 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},\\    
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks which is the relevant case for passively irradiated
PPDs \citep{chiang97}.

\subsection{Vertical shear}\label{vshear_def}
Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot} and Eq. \ref{vertiso_eqm}, but is also 
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}.\label{vertical_shear}
\end{align}
For the above disk models,
\begin{align}
  r\frac{\p\Omega^2}{\p z} = \frac{sz\Omega_k^2}{\Gamma
    r\left(1+z^2/r^2\right)^{3/2}}. \label{vertical_shear_ex} 
\end{align}
Note that there is a maximum vertical shear rate at $|z|=r/\sqrt{2}$,
which is expected to limit the growth rate of instabilities associated with vertical shear.

@@

\subsection{Stability against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $j=r^2\Omega$ is the specific angular
momentum. For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is generally satisfied. 
Thus, to assess stability, we consider the second
criterion. 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\Omega_k^2$ but
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters the right hand
side (RHS) is $O(\epsilon^2)$. Hence, instability in the adiabatic limit  
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for passively irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
thermal relaxation, as shown below.

>>>>

\subsection{Thermal relaxation}
We model radiative effects as thermal
relaxation with a timescale $t_c$, i.e.,
\begin{align}\label{thermal_relax}
  \Lambda  \equiv \frac{\rho\mathcal{R}}{\mu}\frac{T - T_0}{t_c},
\end{align}
where $T_0=T_0(r,z)$ is the equilibrium temperature.  We 
define the dimensionless cooling time 
\begin{align}\label{beta_def}
  \beta \equiv \OmK t_c
\end{align}
relative to the Keplerian frequency, $\OmK=\sqrt{GM_*/r^3}$.

@@

We consider disk equilibria in which the pressure $P$ and density
$\rho$ are related by   
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a  constant parameter  and the
function $c_{s0}(r)$ is prescribed below. Choosing $\Gamma=1$ gives  a 
locally vertically isothermal disk and $c_{s0}(r)$ is the mid-plane
sound-speed. 
It shall be convenient to define the function $c_s(r,z)$ 
\begin{align}
  c_s^2\equiv % \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
  \frac{\Gamma P}{\rho}, 
\end{align}
which is related to the isothermal sound-speed $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$. We take $c_{s0}(r)$ to be another power-law:  
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index, and
define   
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\OmK$ is the Keplerian azimuthal velocity and
$H$ is a characteristic scale-height.  Thus, $\epsilon$
is also a characteristic disk aspect-ratio. We are 
interested in thin disks with $\epsilon \ll 1$.%,%  as relevant for
% protoplanetary disks \citep{chiang97}.

@@

Similarly, the equilibrium rotation is 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\OmK^2(r)}=1 +
  \left(p+\frac{s}{\Gamma}\right)\epsilon^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
where $s\equiv q+p(1-\Gamma)$. We also define % the radial shear rate  
% \begin{align}
%   S(r,z) \equiv r\frac{\p\Omega}{\p r},  
% \end{align}
% and
the square of the epicyclic frequency $\kappa$ 
\begin{align}
  \kappa^2(r,z) \equiv \frac{1}{r^3}\frac{\p }{\p
    r}\left(r^4\Omega^2\right). %2\Omega(2\Omega + S). 
\end{align}

\subsubsection{Vertically isothermal disks}
For $\Gamma=1$ the equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ are given by 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},\\    
  \frac{\Omega^2(r,z)}{\OmK^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks which is the relevant case for passively irradiated
PPDs \citep{chiang97}.

\subsection{Vertical shear}\label{vshear_def}
Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot} and Eq. \ref{vertiso_eqm}, but is also 
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}.\label{vertical_shear}
\end{align}
For the above disk models,
\begin{align}
  r\frac{\p\Omega^2}{\p z} = \frac{sz\OmK^2}{\Gamma
    r\left(1+z^2/r^2\right)^{3/2}}. \label{vertical_shear_ex} 
\end{align}
Note that there is a maximum vertical shear rate at $|z|=r/\sqrt{2}$,
which is expected to limit the growth rate of instabilities associated with vertical shear.

@@

\subsection{Stability against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $j=r^2\Omega$ is the specific angular
momentum. For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is generally satisfied. 
Thus, to assess stability, we consider the second
criterion. 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\OmK^2$ but
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters the right hand
side (RHS) is $O(\epsilon^2)$. Hence, instability in the adiabatic limit  
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for passively irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
thermal relaxation, as shown below.
